\section{Introduction}
\ifsummaries
\begin{itemize}
 \item Blender is a 3D modeling software that also includes a game engine (the Blender Game Engine)
 \item Character animations are known to be slow, which this project attempts to address
\end{itemize}
\fi

Blender is a free and open-source 3D modeling and animation program\cite{blender}.
One of its many features is a built-in game engine called The Blender Game Engine (BGE).
Character animations in the BGE are considered slow by the BGE community.
For comparison, the benchmark scene described in section~\ref{sec:bench_scene} runs at approximately 187ms per frame on the benchmark machine described in section~\ref{sec:bench_platform}.
As a quick test, a scene similar to the benchmark scene was created and ran using and open-source graphics engine called OGRE\cite{ogre}.
In OGRE, the scene ran at approximately 37ms per frame.
Not only is this approximately five times faster, but the OGRE test could probably be further optimized by someone more knowledgeable with OGRE.
The OGRE example uses just the basics needed to get the scene to run.
For example, different scenegraphs, instancing or hardware skinning could be explored and used.
\todo{Add code for the OGRE example (probably as an appendix).}

This work attempts to improve the performance of these skeletal mesh animations in the BGE.
Several optimizations are used in this work, which can be grouped into three main categories: parallelizing code, moving code to the GPU, and other minor optimizations.
